---
title: "Assignment 3"
author: "Xiaoshu Gui"
output: pdf_document
geometry: margin = 1.5cm
---
## Setup
```{r message=FALSE, warning=FALSE}
library(dplyr)
library(purrr)
library(magrittr)
library(tidyr)
library(tibble)
library(stringr)
library(reshape2)
library(lme4)
library(mfx)
library(bayesm)
library(mclogit)
library(mlogit)
library(nnet)
# load data
data(margarine)
# Create a dataframe that merges product characteristics with household demos by hhid.
choiceprice <- as.matrix(margarine$choicePrice)
demos <- as.matrix(margarine$demos)
marg <- merge(choiceprice, demos, by = "hhid")
```
## Exercise 1 Data Description  
* Average and dispersion in product characteristics.
```{r}
# average:
apply(marg[, 3:12], 2, mean)
# dispersion
apply(marg[, 3:12], 2, sd)
```
* Market share, and market share by product characteristics.
```{r}
# market share by product
ms_product <- table(marg$choice)/4470
names(ms_product) <- names(marg[,3:12])
print(ms_product)
# market share by product characteristics: brand and type
brand_name <- names(marg[,3:12]) %>%
  str_replace_all("_Stk|_Tub", "")
ms_brand <- cbind.data.frame(brand_name, ms_product) %>%
  group_by(brand_name) %>%
  summarise(market_share = sum(Freq))
print(ms_brand)
# by product type (stick and tub)
sum(ms_product[1:6]) # market share of stick
sum(ms_product[7:10]) # market share of tub
```
* Mapping between observed attributes and choices.

Create tables of choices by different household attributes:
```{r}
# income level & choices
table(marg$Income, marg$choice)
# family size: 
table(marg$Fam_Size, marg$choice)
# education status & choices
table(marg$college, marg$choice)
# job status & choicies
table(marg$whtcollar, marg$choice)
# retirement status & choices
table(marg$retired, marg$choice)
```
## Recap Multinomial Models
There are *m* alternatives and the dependent variable *y* is defined to take value *j* if the *j*th alternative is taken, j = 1, ..., m. Based on the random utility model, let $U_{ij}$ denote the utility of individual *i* derive when choosing altertive *j*. 
*j* is chosen if and only if $U_{ij} > U_{ik}$ for all $k \neq j$. Although we can't observe $U_{ij}$, we can treat it as independent random variables with a systematic component $V_{ij}$ and a random component $\epsilon_{ij}$ such that $U_{ij} = V_{ij} + \epsilon_{ij}$. 

Define the probability that alternative *j* is chosen by individual *i* as: 
$$P_{ij} = Pr[y_i = j] = \dfrac{V_{ij}}{\sum_{k=1}^{m}V_{ik}}, ~~~  j = 1, ..., m, $$ 
where $V_{ij} > 0$ can be general functions of regressors $X_i$ and parameters $\beta$. This is a *universal logit model*. Different specifications for $V_{ij}$ corresponds to specific models, such as multinomial logit and conditional logit models. In that sense, all these models are variants of the same model. They only differ in their parametrization of the systematic compotents $V_{ij}$.  

The log likelihood function of the universal logit model is (assuming independent realizations by summing up all N individual contributions):
$$ L = \sum_{i = 1}^{N} \sum_{j=1}^{m} y_{ij} ~ ln~ P_{ij}, ~~~~~ j = 1, ..., m, ~~~  i = 1, ..., N,$$
where $P_{ij}$ is defined above. 

## Exercise 2 First Model
* We are interested in the effect of price on demand. Propose a model specification.

Since the price of a product varies by different choices, a conditional logit model is chosen to deal with regressors varying across alternatives. Here $V_{ij} = X_{ij}$, specifying characteristics of the alternatives (price). 

The probability of the *i*th househould choosing product *j* is given by
$$ P_{ij} = Pr[y_i = j] = \dfrac{\exp(X_{ij}\beta)}{\sum_{k=1}^{m}exp(X_{ik}\beta)},~~~ j = 1, ...m $$ 
where $X$ denotes price, the substrcipt *i* denotes the *i*th household, subscript *j* or *k* denotes the alternative, and parameter $\beta$ is contant across alternatives. Note that it is possible to go from alternative-varying regressors to alternative-invariant format. Let $X_i$ be a $K*1$ vector. Define $X_{ij}$ to be a $Km*1$ vector with zeros except that the *j*th block is $X_i$, that is $X_{ij} = [0'... 0', X_i, 0', ...0']'$, and define $$ \beta = [0', \beta_2',... \beta_m']',$$ where $\beta_1 = 0$ is a normalization. Then $X_i'\beta_j = X_{ij}'\beta$. 

The likelihood function of conditional multinomial model is: 
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
loglik_cl <- function(beta) {
  X = marg[, 3:12] # price takes dif values for dif alternatives (1*10)
  X_beta = X * beta # beta is a constant for each decision maker i. (1*10)
  X_beta_j = matrix(nrow = nrow(marg), ncol = 1) # coefficient for a given choice j (N*1)
  for (i in 1:nrow(marg)) {
    jstar = marg[i, "choice"]
    X_beta_j[i] = X_beta[i, jstar] 
  }
  numerator = exp(X_beta_j)
  denominator = rowSums(exp(X_beta))
  pij = numerator/denominator
  ll = log(pij)
  loglik_cl <- -sum(ll)
}
fit_cl = nlm(f = loglik_cl, p = 0)
fit_cl$estimate
```
Interpretation: Note that the estimated beta < 0, suggesting that an increase in the price of one alternative decreases the probability of choosing that alternative and increases the probability of choosing other alternatives. 

## Exercise 3 Second Model
We are interested in the effect of family income on demand. Propose a model specification:

Since family income is a fixed constant for decision makers and does not vary across product choices, a multinomial logit model is chosen to address alternative-invariant regressors. The probability of the *i*th household choosing product *j* is given by:
$$ P_{ij} = Pr[y_i = j] = \dfrac{\exp(\alpha_j + X_{i}\beta_j)}{\sum_{k=1}^{m}exp(\alpha_k + X_{i}\beta_k)}, ~~~j = 1, ...m, $$
where *X* denotes income. The likelihood function is:
```{r}
a <- function(beta){
  X = as.matrix(marg[, "Income"]) # income is a N*1 vector
  beta = matrix(nrow = 1, ncol = 10) # alternative-specific constant (1*10)
  beta[1] = 0  # set beta_1 to 0-- use product 1 as reference group 
  beta_j = matrix(nrow = 1, ncol = 1)
  X_beta = X %*% beta # returns a matrix of N*m
  X_beta_j = matrix(nrow = nrow(marg), ncol = 1) # N*1
 # alpha_t = matrix(rep(t(alpha), times = nrow(data)), ncol = ncol(t(alpha)), byrow = T)
  for (i in 1: nrow(marg)) {
    jstar  = marg[i, "choice"]
    beta_j = beta[jstar]
    X_beta_j[i] = X_beta[i, jstar]
  }
  numerator = exp(X_beta_j)
  denominator = sum(exp(X_beta_j))
  pr = numerator / denominator
  ll = log(pr)
  a = -sum(ll)
}
fit_a = nlm(f = a, p = c(rep(0, 10))) # Fail to debug.

loglik_mnl = function(beta) {
  X = as.matrix(cbind(marg[, 13:15], marg[, 17:19], rep(1, nrow(marg)))) # 4470*7
  beta = matrix(nrow = 7, byrow = T) # create a matrix of 7*1
  X_beta_j = X %*% beta # 4470*1 
  
  for (i in 1:nrow(marg)) {
    jstar = marg[i, "choice"]
    X_beta_j[i] = X[i, jstar] %>% beta
  }
  numerator = exp(X_beta_j)
  denominator = sum(exp(X_beta_j))
  Pij = numerator/denominatorn
  ll = log(Pij)
  loglik_mnl = - sum(ll)
}
# optimize the likelihood of multinomial model:
#fit_mnl = nlm(f = loglik_mnl, p = rep(0, 7)) 
#fit_mnl
```
## Exercise 4 Marginal Effects
Compute and interpret the marginal effects for the first and second models.
```{r}
## Marginal effect of the conditional logit model
X = marg[, 3:12] # N*m
b = fit_cl$estimate # 1
X_beta = X * b # N*m
X_beta_j = matrix(nrow = nrow(marg), ncol = 1) # N*1
xbetak =  exp(X_beta)
denominator = rowSums(xbetak)
pr_ij = as.matrix(xbetak/denominator) # N*m
pij = t(pr_ij) %*% pr_ij * (-b) # m*m (10*10)
margin = matrix(rep(colSums(pr_ij) * b, 10), ncol=10 )
margin = margin * diag(10)
me_cl = (pij + margin)/nrow(marg)
me_cl

## Marginal effect of the multinomial model
X = as.matrix(cbind(marg[, 13:15], marg[, 17:19], rep(1, nrow(marg))))
#beta = matrix(fit_mnl$estimate, nrow = 7, byrow = T)
#X_beta_j = X %*% beta
#ex = exp(X_beta_j)
#Pij = t((apply(ex, 1, function(x) x / sum(x))))
#beta_income = matrix(beta[1, ])
#beta_bar = Pij %*% beta_income 
#beta_bar_large = matrix(rep(beta_bar, 10), ncol = 10)
#beta_j = matrix(rep(t(beta_income)), nrow(data), byrow = T, ncol = 10)
#me_mnl = data.frame(colSums(Pij * (beta_j - beta_bar_large))/nrow(data)) 
#me_mnl
```
## Exercise 5 IIA
Now combine the above two models to estimate the effect of price and family income on choices of margarine. The mixed logit model is specified as:
$$ P_{ij} = \dfrac{exp(X_{ij}\beta ~ + W_i \gamma_j)}{\sum_{k = 1}^{m} exp(X_{ik}\beta + W_i\gamma_k)}, ~~~ j = 1, ..., m$$
Its likelihood function is:
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
loglik_mixed = function(beta) {
  X = marg[, 3:12] - marg[, 3] # set price of the first product as reference 
  b = beta[1] # alternative-variant coefficient
  alpha = beta[2:11] # alternative-invariant coefficient
  alpha[1] = 0 
  X_beta = X * b 
  alpha_choice = matrix(nrow = nrow(marg), ncol = 1)
  X_beta_j = matrix(nrow = nrow(marg), ncol = 1)
  alpha_t = matrix(rep(t(alpha), times = nrow(marg)), ncol = ncol(t(alpha)), byrow = T)
  for (i in 1: nrow(marg)) {
    jstar  = marg[i, "choice"]
    alpha_j = alpha[jstar]
    alpha_choice[i] = alpha_j
    X_beta_j[i] = X_beta[i, jstar]
  }
  numerator = exp(X_beta_j + alpha_choice)
  Xbeta_k =  exp(X_beta + alpha_t)
  denominator = rowSums(Xbeta_k)
  Pij = numerator / denominator
  ll = log(Pij)
  loglik_mixed = - sum(ll)
}
# optimize the likelihood using the nlm function. 
fit_mixed = nlm(f = loglik_mixed, p = c(rep(0, 11)))
beta_f <- fit_mixed$estimate
```
# Recap: The IIA assumption

The ratio of logit probabilities of any two alternatives *j* and *k* is
$$ \dfrac{Pr(y_i = i)}{Pr(y_i = k)} = \dfrac{exp(V_{ij})}{exp(V_{ik})} = exp(V_{ij}- V_{ik})$$
Note that the above ratio only depends on alternatives j and k. Because the ratio is independent of alternatives other than j and k, MNL logit models are said to be independent of irrelevant alternatives (IIA). IIA implies that presence or absence of another alternative should not alter the relative probabilities of any single decision-maker (conditional on the model’s systematic
component).

* Consider an alternative specification, where we remove one choice (the last one) from the data. 
```{r}
sub_marg <- marg %>% # create a subset of data which remove the 10th choice.
  filter(choice < 10)
loglik_mixed_2 = function(beta) {
  X = sub_marg[, 3:11] - sub_marg[, 3] # do not include the price of 10th product
  b = beta[1] # alternative-variant coefficient
  alpha = beta[2:10] # alternative-invariant coefficient
  alpha[1] = 0 # set the first product as reference group
  X_beta = X * b 
  alpha_choice = matrix(nrow = nrow(sub_marg), ncol = 1)
  X_beta_j = matrix(nrow = nrow(sub_marg), ncol = 1)
  alpha_t = matrix(rep(t(alpha), times = nrow(sub_marg)), ncol = ncol(t(alpha)), byrow = T)
  for (i in 1: nrow(sub_marg)) {
    jstar  = sub_marg[i, "choice"] 
    alpha_j = alpha[jstar]
    alpha_choice[i] = alpha_j
    X_beta_j[i] = X_beta[i, jstar]
  }
  numerator = exp(X_beta_j + alpha_choice)
  Xbeta_k =  exp(X_beta + alpha_t)
  denominator = rowSums(Xbeta_k)
  Pij = numerator / denominator
  ll = log(Pij)
  loglik_mixed_2 = - sum(ll)
}
fit_mixed_2 = nlm(f = loglik_mixed_2, p = c(rep(0, 10)))
beta_r <- fit_mixed_2$estimate
```
By dropping one alternative (further tests can drop more irrelavant alternatives at a time) and reestimating the model, we can see that the coefficients do not change, indicating that IIA might hold. 

* Compute the test statistics:

Since
$$ MTT = -2[L_r(\beta^r)- L_r(\beta^f)] = -2 ln \dfrac{L(\beta^r)}{L(\beta_f)} = 2 ln \dfrac{L(\beta^f)}{L(\beta_r)} = LR,$$
where $ L(\beta^f)$ is the likelihood evaluated at the MLE and $ L(\beta^r)$ is the maximum of likelihood subject to the restriction (that r parameters unconstrained in the full likelihood analysis are assigned fixed values). 
```{r}
# calculate the likelihood evaluated at MLE (beta_f)
X = marg[, 3:12] - marg[, 3] 
b = beta_f[1] 
alpha = as.matrix(beta_f[2:11]) # 10*1
X_beta = X * b 
alpha_t = matrix(rep(t(alpha), times = nrow(marg)), ncol = ncol(t(alpha)), byrow = T) # 4470*10
Xbeta_k =  exp(X_beta + alpha_t) # 10*10
denominator = rowSums(Xbeta_k)
Pij = Xbeta_k / denominator
ll = log(Pij)
loglik_beta_f = sum(ll)
# ML subject to restriction (beta_r)
X = sub_marg[, 3:11] - sub_marg[, 3] 
b = beta_r[1]
alpha = beta_r[2:10] 
X_beta = X * b 
alpha_t = matrix(rep(t(alpha), times = nrow(sub_marg)), ncol = ncol(t(alpha)), byrow = T)
Xbeta_k =  exp(X_beta + alpha_t)
denominator = rowSums(Xbeta_k)
Pij = Xbeta_k  / denominator
ll = log(Pij)
loglik_beta_r = sum(ll)
## compute the test statistics
mtt <- log(loglik_beta_f/loglik_beta_r) *2 
# For sufficiently large sample size, the LR test statistic is chisqured distributed
# a χ2 with r degrees of freedom
pchisq(mtt, df = length(beta_r))
```
Another way to calculate the Hausman and McFadden Test by hand:
```{r}
beta_f1 <- beta_f[1:10]
beta_diff <- beta_r - beta_f1
hm <- beta_diff %*% solve(var(beta_r) - var(beta_f1)) %*% t(beta_diff)
pv <- pchisq(hm,df = 2*10) # the degrees of freedom of the Chi-Square distribution used to test the LR Chi-Square statistic is defined by the number of models estimated (2) times the number of predictors in the model (10)
```
The result of statistical tests suggests that IIA holds. 